---
title: "Types of Data"
output: ioslides_presentation
---

## An analogy from math

In math we have various types of numbers. familiar ones are:

* The integers: $\dots, -2, -1, 0, 1, 2, \dots$
* The rational numbers: $p/q$ where both $p$ and $q$ are integers, $q \neq 0$.
* the real numbers: all numbers between $(-\infty, \infty)$
* The infinite numbers such as $\infty$.
* the complex numbers: all values $x + yi$ where $x$ and $y$ are real numbers and $i=\sqrt{-1}$.

----

In math, we can recongnize the needs of such numbers as they come into play when trying to solve equations or do operations:

* Rational numbers are needed to solve something like $5x = 2$.
* real numbers are needed to solve something like $x^2 = 2$
* infinite numbers are needed to discuss arbitrarily large 
* complex numbers are needed to solve something like $x^2 + 2 =0$.

## Numbers in R

With computers we have a primary numeric type: *floating point*

Floating point numbers are a [leaky abstraction](http://www.johndcook.com/blog/2009/04/06/numbers-are-a-leaky-abstraction/) of the real numbers.

Floating point numbers are represented in binary with three values:

* a sign (plus or minus)
* a number (significand) interpreted as a value between $[1,2)$
* an exponent

These are put together as `-1.11001 * 2^(3)`.


----

Why are is this leaky:

* the significand can only represent so many different values. In 64-bit, there are $2^{53}$, or about:

```{r}
2^53
```

A lot -- but not infinite. This means there are "gaps" between the numbers that can be represented. Near 1, these gaps are about $10^{-16}$.

----

These gaps can appear in simple statements:

```{r}
(.1 + .2) - .3
```

* The exponent can only be so big. But big enough to represent numbers as bit as $10^{300}$ and as small as $10^{-300}$, so pretty big and small.

----

In R, most all numbers created are floating point values internally. How they display can make them look like integers:

```{r}
1.00000000000000
```

R assigns a class to these numbers, `numeric`:

```{r}
class(1)
```

This is not the storage type though, which is found with

```{r}
typeof(1)
```


----

R has a type for integers and way to create them, but they are not widely used outside of programming purposes:

```{r}
class(1L)
```

Integers are exact, but limited. They can have faster math defined for them and have some other advantages, but that isn't our concern.

----

R has no built-in type for rational numbers. Rationals, are just floating point numbers:

```{r}
1/2
```

This means rational numbers aren't *exact* either

```{r}
1/10 + 2/10 - 3/10
```

-----

R has a means to represent complex numbers:

```{r}
2 + 3i  # also more cumbersomely: complex(1, 2, 3)
```

-----

If all we wanted to discuss here was numeric data, we would not have much more to say about types. Unless your data is unusual, R will store it as floating point numbers. (Well, not really. When R interfaces with other systems it must be able to interact with a myriad of different representations.)

The mathematical operations are predictable:

```{r}
1 + 2 * 3 / 4 ^ 5
```

Basic functions are called as they would appear in a math book:

```
sin(pi/2)
```

----

The value `pi` is one of a few built-in constants. The number `e` is not one of them. For that, the function `exp(1)` can be used.

Not mentioned, but part of floating point, are the value plus or minus infinity: `Inf` (an answer to `1/0`) and `NaN` for "not a number" which is a way to signal an error without stopping a program. (Generated, for example, by `0/0`.)

## Logical values

In computer programming, Boolean values are commonly used. R has the values `TRUE` and `FALSE`.

```{r}
TRUE
```

These are often generated as answers to questions comprised of logical operators:

```{r}
x  = 1
x > 0
```

----

The "algebra" of Boolean values is implemented by `&` (and) and `|` (or):

```{r}
TRUE & FALSE
TRUE | FALSE
```


## Dates and times

R has built-in classes for storing dates and times. As well, their are other external packages, such as `lubridate`, to facilitate this.

The `POSIXct` class represents the (signed) number of seconds since the beginning of 1970 (in the UTC time zone) as a number vector (`?POSIXct`):

```{r}
now <- Sys.time()
now
```

----

Arithmetic is possible, as with:

```{r}
now - 60*60 # an hour ago
```

----

The `POSIXlt` class stores values in list. It may print the same way, but it allows some underlying access:

```{r}
a <- as.POSIXlt(now)
a
```

```{r}
a$mday  # day of month
```

Time is pretty complicated when timezones, leap seconds, leap years, ... are included. The manual page is informative


## Lubridate

The [lubridate](https://github.com/hadley/lubridate) package can make working with time data fairly intuitive. Assuming it is installed, we load it with:

```{r}
require(lubridate)
```

With this we can create time/date values easily enough:

```{r}
now()
```
----

Or using one of the constructors, such as `ymd` or `ymd_hms`

```{r}
today <- ymd("2016-02-19")
today
today_start <- ymd_hms("2016-02-19 13:00:00", tz = "EST") # OlsonNames()
today_start
```

----

These constructors are smart about leap years:

```{r}
ymd("2016-02-29")
ymd("2017-02-29")
```

----

We can manipulate times with natural idioms:

```{r}
today_start + weeks(1)  # next week's start 
```

There is more to say, but not here....

## Character values

Historically data in the social sciences was classified as one of "nominal", "ordinal", "interval", or "ratio". The distinctions can be read [here](https://en.wikipedia.org/wiki/Level_of_measurement). 

For "nominal", we have

> The nominal type differentiates between items or subjects based only on their names or (meta-)categories and other qualitative classifications they belong to; thus dichotomous data involves the construction of classifications as well as the classification of items. Discovery of an exception to a classification can be viewed as progress. Numbers may be used to represent the variables but the numbers do not have numerical value or relationship.

----

We do see numbers representing such categories frequently. The Census bureau values typically encode different categorical responses using numbers.

More human-friendly is to use names for such data. R has two different ways to store such data, this distinction might help:

* descriptive data is best stored in free-form test
* categorizing data is best stored in a limited set of options. 

When enterind data in excel, we see this in action with auto-fill turned on.

----

Character data in R is formed by using matching single or double quotes.

```{r}
qwerty <- "the quick brown fox jumped over the lazy dog"
lincoln <- "four score and seven years ago..."
````

Character data can be encoded to accommodate other languages. This topic can be involved, but is necessary for many uses. Here we assume ASCII encoding are used.

----

There are a number of functions that manipulate character data:

```{r}
toupper(qwerty)  # also tolower
chartr("q", "Q", qwerty) # more fine-grained translation (never used before!)
substr(qwerty, 5,9)  # take substrings
```

----

Strings can be split up with `strsplit`:

```
strsplit(qwerty, " ")
```

(The otupt is a bit confusing for now.)

----

Strings can be combined. It is a bit cumbersome in R compared to other languages. The `paste` command can be used:

```{r}
paste("the", "quick", "brown", "fox")
```

`paste` has an argument for what `sep`arates the values, with a default of `" "`. The `paste0` function is identical with default separation of `""`.

----

R has pattern matching through regular expressions available. A useful function is `grepl` that indicates if a match occurs:

```
grepl("QUICK", qwerty)  # case sensitive
sapply(letters, function(x) grepl(x, qwerty))  # all there...
```

(This last command hopefully will make sense by the end of these notes...)


## Factors

For categorical data, there are a *fixed*  number of levels from which a value comes from. There are computational advantages here, as storage can be made more efficient. There are other statistical advantages too.

Putting that asside, such values are called "factors" and are created by the `factor` function. Let's make the Likert scale. (The levels are generated with a vector which will be explained in a bit):

----

```{r}
likert <- c("strongly agree", "agree", "no opinion", 
            "disagree", "strongl disagree")
x = factor("agree", levels = likert)
x
```

The value of `x` is secretly an integer:

```{r}
storage.mode(x)
```

This can cause confusion, as the levels might look numeric, but use a different ordering.

----

We can't set a value that is not one of the levels

```{r}
factor("AGREE", levels=likert)
```

So spelling matters.

----

Oops, I spelled "strongly" as `strongl`. How to change that?

We could correct the values of `likert` and then re-create `x`. But we may not want to do that. We can also redefine the levels of a factor. Here is one way:

```{r}
likert <- c("strongly agree", "agree", "no opinion", 
            "disagree", "strongly disagree")
levels(x) <- likert
x
```

> For those with experience programming, the left side with `levels(x)...` should seem a bit odd. This is R's syntax for mutating the value that `x` is assigned to and is idiosyncratic.

---- 

Levels can also be collapsed and recoded. The syntax below uses a list, to be described later. For now it consists of a pairs of new labels and old labels to collaspe:

```{r}
levels(x) <- list("AGREE"    = c("strong agree", "agree"), 
                  "BLAND"    = c("no opinion"), 
                  "DISAGREE" = c("disagree", "strongly disagree")
)
```

----

If you think this all seems like a pain, you aren't alone. However, by default, R uses factors where you might want simple character data. So, knowing what is happening is important.

## Missing Data

We saw that when we tried to create a factor from a value that wasn't a level, we got an odd answer:

```{r}
factor("AGREE", levels=likert)
```

What exactly is `<NA>`? In this case the proposed value isn't one of the agreed upon levels, so the actual answer is missing. In R the value `NA` stands for "Not available" or "missing." A value can be unknown, hence missing; or unknowable, and missing.

```{r}
NA
```

----

An interesting think about `NA` values is that they poison subsequent operations:

```{r}
5 + NA
```

Some functions have arguments that instruct the computation to proceed after dropping `NA` values.

----

How might we identify an `NA` value? We could try this:

```{r}
NA == NA
```

Hmmm, why might that not work?

The `is.na` function is here for this:

```{r}
is.na(NA)
```

(This is different, but somehow related to, the floating point value `NaN`.)

## Coercion between types

R has numerous "`as`" functions to coerce one data type into another.

Earlier, we slipped in this conversion:

```{r}
as.POSIXlt(Sys.time())
```

----

For another example, we can make a string numeric (if sensible) through:

```{r}
as.numeric("123")
as.numeric("-1.23e4")
as.numeric("one-two-three")   # fails
```

----

We can find the level of a factor through `as.numeric`:

```{r}
x <- factor("agree", levels=likert)
as.numeric(x)
```

But what if our factor is labeled by numbers:

```{r}
x <- factor(5, c(5,10,15,20))
x
as.numeric(x)
```

-----

Okay, how do we get the value `5`? First by coercing to a character and then numeric:

```{r}
as.numeric( as.character( x ) )
```

This is a somewhat common thing to do. R makes some things easy though. Even though this isn't quite right, equality will convert things quietly to make this test work as desired:

```
x == 5
```

----

More generally, to coerce an object of one class into another, we have the `as(obj, classname)` function:

```{r}
as(1.0, "integer")
```

## Questions

XXX

